---
title: "Preparing micro-array data for coGSEA analysis"
output: html_notebook
---

## Reading and preparing data for [coGSEA](link) analysis.

In this document, we will show you how to prepare micro-array data to perform a coGSEA analysis.  

This example dataset is from a [article](https://www.ncbi.nlm.nih.gov/pubmed/21836821) about Astrocymotas tumors published in 2011 by Liu et al.  

You can download it on the Gene Expression Omnibus database with the accession number [GSE19728](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE19728)  

#

#### Loading the necessary packages for this analysis

```{r, eval}
library(affy)
library(hgu133plus2.db)
library(edgeR)
library(gage)
library(coGSEA)
```

#### Loading the CEL FILES and Normalizing them usin `rma`

```{r}
setwd("~/GitLab/GSE19728/data/")
celfiles = ReadAffy()
celfiles = rma(celfiles)
```

#### Getting the expression data, and the Probe Names

```{r}

intensity = exprs(celfiles)
intensity = cbind(rownames(intensity), intensity)
colnames(intensity)[1] = "PROBEID"
intensity = as.data.frame(intensity)
intensity$PROBEID = as.character(intensity$PROBEID)
```

#### Getting the annotations of Probes IDS to ENTREZ accession numbers

```{r}
annots = select(hgu133plus2.db, intensity$PROBEID, "ENTREZID", "PROBEID")
```

#### Merge expression matrix and annotation

```{r}
res = merge(intensity, annots, by= "PROBEID")

```


#### Getting rid of PROBE ids and type casting 

```{r}
resmin = res[,2:ncol(res)]
cname = colnames(resmin)
resmin = apply(resmin, 2, as.numeric)
colnames(resmin)= cname
resmin = as.data.frame(resmin)
```

#### Aggregating the PROBES matching the same ENTREZ accession number by averaging them and applying a log transformation

```{r}
result = aggregate(. ~ ENTREZID, resmin, mean)
result$ENTREZID = levels(as.factor(as.character(resmin$ENTREZID)))
rownames(result) = result$ENTREZID
result = result[,-1]
result = log(result)
```

#### Visualizing in boxplot to check normalization

```{r}
boxplot(result, las = 2)
```

#### Changing column names to remove the `.CEL` filename extension

```{r}
colnames(result) = gsub(".CEL","", colnames(result))
print(colnames(result))
```


#### Selecting only the samples we are interested in

```{r}
result2 = cbind(result$GSM492649_Astrocytomas_N, result$GSM525014, result$GSM525015, result$GSM525016, result$`GSM492662_Astrocytomas_T4-1`, result$`GSM492663_Astrocytomas_T4-2` , result$`GSM492664_Astrocytomas_T4-3`, result$`GSM492665_Astrocytomas_T4-4`, result$`GSM492666_Astrocytomas_T4-5`)
colnames(result2) = c("GSM492649_Astrocytomas_N", "GSM525014", "GSM525015", "GSM525016","GSM492662_Astrocytomas_T4-1", "GSM492663_Astrocytomas_T4-2", "GSM492664_Astrocytomas_T4-3", "GSM492665_Astrocytomas_T4-4", "GSM492666_Astrocytomas_T4-5" )
rownames(result2) = rownames(result)
```

#### Preparing the design matrix

```{r}
Normal = c(rep(1,4),rep(0,5))
Tumor = c(rep(0,4),rep(1,5))
design = cbind(Normal, Tumor)
rownames(design) = colnames(result2)
```

#### Preparing the contrast matrix

```{r}
contr.matrix = makeContrasts(NormalVSTumor = Normal - Tumor, levels = design)
```


#### Preparing expression list object

```{r}
temp = new("EList")
temp$design = design
temp$E = as.matrix(result2)
rownames(temp$E) = as.numeric(rownames(temp$E))
temp$genes$ENTREZ = rownames(result2)
temp$common.dispersion = estimateDisp(temp$E, design = temp$design)$common.dispersion
temp$samples = colnames(result2)
```

#### Preparing gene set collection

```{r}
gs = gage::kegg.gsets(species = "hsa", id.type = "entrez")
geneset = gs$kg.sets
```

#### Function for simplifying gene sets names

```{r}
nameshorter = function(names){
  namemod = c()
  for (i in seq(1,length(names))){
    namemod[i] = paste(strsplit(names[i], " ")[[1]][-1], sep = "", collapse = " ")
    namemod[i] = gsub("/","", names[i])
    namemod[i] = gsub(" ","_", names[i])
  }
  return(namemod)
}
```

#### Simplifying gene sets names

```{r}
names(geneset) = nameshorter(names(geneset))
names(geneset) = gsub("/","_",names(geneset))
```

#### Saving necessary objects to RDS files

```{r}
saveRDS(contr.matrix, "~/GitLab/GSE19728/contrast.rds")
saveRDS(temp, "~/GitLab/GSE19728/elist.rds")
saveRDS(geneset, "~/GitLab/GSE19728/geneset.rds")
```


#### Reading necessary objects (generated above) from RDS files

```{r}
elist = readRDS("~/GitLab/GSE19728/elist.rds")
contrast = readRDS("~/GitLab/GSE19728/contrast.rds")
geneset = readRDS("~/GitLab/GSE19728/geneset.rds")
```



#### Running coGSEA analysis 

```{r}
coGSEA(ElistObject = elist, contrastMatrix = contrast, ENTREZGenesIds = elist$genes$ENTREZ, geneSetCollection = geneset,specie = "Homo sapiens", directoryPath = "~/GitLab/GSE19728/results", alpha = 0.05, pvalAdjMethod = "BH", pvalCombMethod = "sumlog",min.intersection.size = 1, GSEA.Methods = c("camera", "gage","globaltest", "gsva", "ssgsea", "zscore", "ora", "padog", "roast","safe"), num.workers = 4, shinyMode = FALSE)

```